CrystalPlan: an Experiment Planning Tool for Crystallography

### 2010-12-30 docs/CrystalPlan_paper/paper.tex
Refs #182: Christina's changes. Should be the last draft before submitting.

\begin{abstract}
Beam time at large x-ray and neutron scattering facilities is always at a premium.
The CrystalPlan program can calculate the data coverage of a crystal in reciprocal
space in a single-crystal diffraction time-of-flight experiment. CrystalPlan can 
help a user build an experiment plan that will acquire the most data possible, 
with sufficient coverage but limited redundancy, therefore increasing scientific 
productivity. 
A user friendly GUI including a 3D viewer, an automated coverage optimizer,
and an option to reorient the crystal for the measurement 
of selected hkls on specific detector positions 
are among its useful features. 
A sample use case of the program with the TOPAZ beamline at SNS will be
presented. 
\end{abstract}

@@

The TOPAZ beamline at the Spallation Neutron Source (SNS) at Oak Ridge National
Lab (ORNL), currently finishing commissioning, is a neutron time-of-flight
Laue instrument with high wavelength resolution. It is designed to acquire
single-crystal diffraction data of small-to-moderate sized unit cells with  high
throughput. The final design calls for an array of 48 detectors covering a large
fraction of $2\pi$ steradians. However as of this writing, only 14 of the 48
detectors are installed, giving only a fraction of $2\pi$ steradian coverage. 
In order to have sufficient data to perform a fit, an experimenter will typically want
to measure $ > 85\%$ of the peaks within a certain range of d-spacings. 
However, wide wavelength band combined with TOPAZ's complex
geometry makes it very difficult to come up with an effective set of sample
orientations that will cover all the required peaks without unnecessary redundancy.

Existing diffraction experiment planning software programs (for example,
CrysAlis Pro from Oxford Diffraction; COSMO from Bruker AXS) were designed for
X-ray experiments where a single  incident wavelength and a large number of
sample orientations or sweeps are used to collect data. 
These programs are not suitable for TOPAZ, since as a time-of-flight instrument, 
it is able to collect data from a wide bandwidth of
incident neutrons, using a stationary crystal. 
Additionally, this means that relatively few sample orientations are needed in
order to complete a measurement: whereas an X-ray measurement might have
hundreds or thousands of individual frames, a TOPAZ run may have as few as 10.

The goal of CrystalPlan was to make an easy-to-use software tool to aid users in
planning their experiments. By quickly making an experimental plan that will be
sufficient for the needs of the scientist (enough peaks measured), but without
spending time needlessly measuring the same region in reciprocal space,
CrystalPlan will greatly increase the scientific productivity and throughput of
TOPAZ and other beamlines. Users will be able to measure more
samples or more fully explore parameters such as temperature and pressure.

@@

\begin{itemize}
  \item Load an instrument geometry.
  \item Calculate the orientation of a crystal at different goniometer settings.
  \item Calculate the coverage of these orientations in reciprocal space,
  in volume and for single peaks.
  \item Take crystal Bravais lattice symmetries into account.
  \item Optimize the experiment plan automatically.
  \item Compare predicted and actual measurement of peaks.
\end{itemize}

@@

CrystalPlan supports area detectors with flat faces and arbitrary rectangular
dimensions; this handles the detector types in use at SNS (Anger cameras and
8-pack tube detectors), but the software could easily be extended for more
complicated detector shapes as needed. The detector configuration is loaded from
a text file containing the location, size, orientation and number of pixels of
each detector. Other settings to be set for an instrument include the
detectable range of neutron/x-ray wavelength.

@@

CrystalPlan is designed to be able to use a variety of sample orientation
goniometers. A base Goniometer class is subclassed for specific types of
goniometers, allowing a unified interface. This base class acts as a
\emph{universal goniometer},
which uses the conventional goniometer angles of
$\phi$, $\chi$, $\omega$ as defined in \cite{busing67}.
Each class inheriting from Goniometer can convert
the desired $\phi$, $\chi$, $\omega$ angles to the internal motor positions it requires. The API allows
for limited goniometers: for example, TOPAZ is currently equipped with a
goniometer where chi is fixed at 45 degrees. CrystalPlan handles this by
limiting the available sample orientation angles accessible to the user.

From an initial data set, the data analysis program ISAW \cite{Tao2006} is used
to calculate the the sample mounting orientation UB matrix, which is saved to a text file along with the lattice
parameters. CrystalPlan can load this UB matrix file and will use that UB matrix
in coverage calculations.

@@

The reciprocal space to be measured is divided into a cubic 3D matrix, the size
of which is determined by the user, who specifies the minimum d-spacing
($d_{min}$, in \ang) that he or she is interested in. The resolution of the
3D matrix is also specified (in \ang$^{-1}$), which determines the number of
points (voxels or volumetric pixels) to be modeled. Available machine memory
limits the resolution that can be achieved in practice.

Each point of the reciprocal space 3D matrix holds an integer representing the 
number of times that voxel of {\bf Q}-space (with {\bf Q} defined as $\frac{2\pi}{d}$) has
been measured -- this is the total coverage matrix. To speed up the
calculation of the total, a coverage matrix for each sample orientation is 
calculated and saved in memory. This means that the total coverage matrix can
be computed by simply adding multiple 3D matrices corresponding to each sample orientation.

@@

For each sample orientation, a rotation matrix combining the sample's mounting
orientation (the U matrix loaded from ISAW) and the goniometer rotation matrices
($\Phi$, $\chi$, $\Omega$) is computed. The coverage in reciprocal space is
computed in this way: for each pixel in the face of a detector, the direction of
the scattered beam is known. Then, two incident beam vectors are considered,
corresponding to the $\lambda_{min}$ and $\lambda_{max}$ wavelength limits
of the detectors and/or source. The scattered beam direction and the two possible incident beam vectors
are used to to find two $\overrightarrow q$ vectors; these point in the
same direction. At the high wavelength we find $\overrightarrow q_{min}$, and
the low wavelength limit scattering off of $\overrightarrow q_{max}$. Since the
instrument has a wide bandwidth, we fill in the 3D {\bf Q}-space matrix by
filling all points from $\overrightarrow q_{min}$ to the 
$\overrightarrow q_{max}$, for all pixels in the face of the detector.
This quickly defines a volume of coverage, as shown schematically in
 two dimensions in Figure \ref{fig:ewald_fig}.

To reduce memory usage yet allow for quickly trying different detector
configurations, the 3D matrix of each sample orientation is saved a 32- or
64-bit integer, with each bit representing a particular detector. A simple
binary mask can then be applied to quickly compute the coverage if some
detectors are disabled, for example.

@@

Each $hkl$ peak is a separate object that holds a list of predicted
measurements, including which detector measured it, at what position on the
detector face, at what wavelength, and for which sample orientation. These data
will be used to  graphically display coverage and calculate statistics such as
redundancy.

\subsection{Crystal Symmetry}

When entering the lattice parameters, the crystal's point group symmetry can
also be entered. This information is then used to calculate the increased
measurement redundancy \cite{giacovazzo92}. A list of $n$ $3\times3$
multiplication matrices is generated based on the point group; any $hkl$ vector can be multiplied by
these matrices to find the $n$ equivalent $hkl$ reflections; for example, for
orthorombic (mmm) symmetry, $n=8$; therefore 8 multiplication matrices are
generated that multiply $hkl$ to give $\pm h$, $\pm k$, $\pm l$.

For the single reflections, each $hkl$ is tracked separately, but a primary
$hkl$ is found that points to a list of equivalent $hkl$'s. The GUI can
then display either all $hkl$, ignoring crystal symmetry; or combine them to
display only part of the reflections. Primary $hkl$'s
are picked so that they make a contiguous block.

In volume coverage mode, a 4D array is generated where the first 3 dimensions
correspond to {\bf Q}-space, and the $4^{th}$ dimension is a list of $n$
entries. Each entry in the $4^{th}$ dimension points to a voxel that is an
equivalent due to crystal symmetry. This allows CrystalPlan to quickly compute a volume coverage
map that takes into account crystal symmetry: at each voxel, all $n$ voxels are
checked to see if any of them were measured.

@@

CrystalPlan allows the user to manually enter lists of sample orientations to
simulate; however, finding an optimal plan by trial and error would be very time
consuming. To make the process easier, CrystalPlan includes an Automatic
Coverage Optimizer, which takes as its inputs the desired number of sample
orientations and uses a genetic algorithm to find a solution that satisfies a
user-specified coverage criterion (for example, the user may ask for $85\%$ of
peaks to be measured, or for $90\%$ of {\bf Q}-volume).

Genetic algorithms have been covered in the literature before, see e.g
\cite{Goldberg89}. In this application, the genes
consist of goniometer angles. For instance, if a goniometer has freedom in 3 angles and the user
requests a 10 different sample orientations, then each individual has 30 genes.
The initial population is created randomly and is made to evolve using mutations
and crossover. Individuals whose goniometer angles turn out to be impossible to
reach are eliminated. For each individual set of sample orientations, the
coverage statistics are computed (taking crystal symmetry into account if desired), and this is used as the fitness of the given individual in the normal genetic
algorithm process.

@@

The program needs to be cross-platform, and run equally well on Linux, Mac OS
and Windows, although Linux is our primary target operating system. Another
design goal was to use of open-source software and toolkits as much as possible.

To this end, CrystalPlan was written in Python \cite{python}, a powerful, open,
cross-platform scripting language. Python was chosen for its ease of cross-platform deployment, and for
its attractive language features. The numpy and scipy libraries (two
open-source, commonly installed Python libraries) were used for many of the
calculations \cite{numpy,scipy}. Some of the critical calculations were
also written with inline C code, which speed up some calculations by up to $400\times$ as compared with
the pure Python version (which can still be used, in case of incompatibilities).

@@

The calculation code of CrystalPlan is placed in a model layer and kept
completely separate from eventual GUI elements, using a Model-View-Controller
architecture \cite{Gamma95}. This makes it possible to run calculations using
Python scripts, if desired. Unit tests for each part of the calculations are used to ensure
consistent results on different platforms and  systems.

@@

An attractive and easy-to-use GUI was considered an important part of
CrystalPlan, since the program is meant to be used interactively as a guide to
the experimenter. The user interface was written using wxPython
\cite{wxPython}, a mature cross-platform GUI toolkit, as well as the Enthought
Traits GUI. Some of the 2D plots use matplotlib \cite{matplotlib},
and the 3D visualization elements use the Mayavi2 toolkit for Python
\cite{mayavi}, itself based on VTK \cite{VTK}. All of these libraries are
open-source.

@@

\begin{itemize}
  	\item {\bf Q}-Space tab: define the volume of interest (minimum d-spacing)
  	and resolution of reciprocal space. 
 	\item Detectors tab: load instrument detector geometry
files, and enable or disable individual detectors. 
	\item Goniometer tab: to choose the
goniometer and degrees of orientation freedom of the instrument.
  	\item Sample tab: enter or load the crystal's lattice parameters and UB
  matrix.
  	\item Try an Orientation tab: interactively rotate a sample and observe the
  changes in coverage in real-time.
  	\item Add Orientations tab: type in a list of sample orientation angles to
  add to the experiment plan.

  	\item Experiment Plan tab: a list of the previously calculated sample
orientations. Here you can manage the orientations you will use in the experiment: add,
delete, or temporarily disable them to see the effect on coverage. Once
complete, the plan can be sent to the SNS data acquisition computers in order to
begin a run \cite{SNSDAS}.

@@

The volume coverage shows a solid isosurface representing the volume measured at
least once. Redundant regions can also be shown using semi-transparent
isosurfaces; or the coverage can be inverted, showing non-measured volumes only;
or sliced between user-selectable $q_{min}$ and $q_{max}$.

@@

Both views can take into crystal symmetry in their display using the techniques
described previously. 
The volume coverage (Figures \ref{fig:volume_view}b) fills in the sphere of
coverage; 
whereas the single reflection view (Figure \ref{fig:refl_view}b) leaves only
the primary reflections (one eight of the total, in this example).

Finally, both views display quick coverage statistics in the lower right
corner, showing the percentage of volume (or of the number of peaks) measured,
as well as the proportion of redundancy (voxels or peaks measured 2 or more
times).

\subsubsection{Single Reflection View.}
Figure \ref{fig:single_reflection} shows a screenshot of the GUI allowing one to
observe in detail the predicted measurements of a particular peak, by clicking
in the 3D view or by typing in the $hkl$ values. The position on the detector
and wavelength of measurement are displayed. An additional window (not shown) allows the user to place the peak on a particular
position on a detector by automatically finding the necessary sample orientation
angles. This is used, for example, to center important peaks in the most
sensitive part of a detector.

@@

The resulting UB matrix was loaded into CrystalPlan, where the Automatic
Coverage Optimizer was used to generate an experiment plan with 8 sample
orientations (Figure \ref{fig:natrolite_gui}a) that covered 97\% of the peaks,
taking into account the crystal symmetry (orthorombic or mmm) (Figure
\ref{fig:natrolite_gui}b). The plan was left to run unattended overnight.

@@

One planned future development for CrystalPlan include extending the software
to help predict coverage in inelastic neutron scattering experiments. Another
desired feature is the introduction of feedback loops to help determine the
right time to end data acquisition at a particular sample orientation.
CrystalPlan would interface with data analysis software such as ISAW and use
recently developped capabilities for real-time data analysis to monitor
the  signal to noise ratio on particular peaks. When the signal becomes
sufficient, the sample will automatically be moved to the next sample
orientation, further minimizing data acquisition time. Combined with advanced
sample changers coming on-line at TOPAZ, this will allow for extremely
efficient, unattended data acquisition.

>>>>

\begin{abstract}
Beam time at large user program based x-ray and neutron scattering facilities is
in high demand and always at a premium. CrystalPlan, a highly efficient
experiment planning software has been developed to maximize the use of available
beamtime per sample per experiment. 
This program can calculate and optimize the data coverage of a crystal in
reciprocal space in a single-crystal diffraction time-of-flight experiment.
CrystalPlan can help a user build an experiment plan that will acquire the most
data possible, with sufficient coverage but limited redundancy, therefore
increasing scientific productivity. 
A user friendly GUI including a 3D viewer, an automated coverage optimizer,
and an option to reorient the crystal for the measurement 
of selected hkls on specific detector positions 
are among its useful features. 
A sample use case of the program with the TOPAZ beamline at SNS will be
presented. 
\end{abstract}

@@

The TOPAZ beamline at the Spallation Neutron Source (SNS) at Oak Ridge National
Laboratory (ORNL), currently finishing commissioning, is a neutron
time-of-flight Laue diffractometer with high wavelength (time) resolution. It
is designed to acquire single-crystal diffraction data of small-to-moderate sized unit cells with  high
throughput. The final design calls for an array of 48 detectors covering three
quarters of $2\pi$ steradians. However as of this writing, only 14 of the 48
detectors are installed, giving only a quarter of $2\pi$ steradian coverage. In
order to have sufficient data for a reliable structure solution and
refinement, an experimenter will typically want to measure $ > 85\%$ of the
peaks within a certain range of d-spacings. However, wide wavelength band
combined with TOPAZ's complex geometry makes it very difficult to come up with
an effective set of sample orientations that will cover all the required peaks
without unnecessary redundancy.

Existing diffraction experiment planning software programs (for example,
CrysAlis Pro from Oxford Diffraction; COSMO from Bruker AXS) were designed for
X-ray experiments where the crystal is rotated stepwise through the Bragg
diffraction condition using a single incident wavelength, 
resulting in a large number of sample orientations.
These programs are not suitable for TOPAZ, since as a time-of-flight instrument
(time-resolved Laue), it is collecting data from a wide
bandwidth of incident neutrons, contracting the Ewald sphere through the
reflection condition thus while keeping a crystal stationary, as described in
[!!!! CC WILSON REFERENCE].
 Additionally, this means that relatively few sample
orientations are needed in order to complete a measurement: whereas an X-ray measurement might have
hundreds or thousands of individual frames, a TOPAZ run may have as few as 10
to cover the same volume of reciprocal space.

The goal of CrystalPlan was to make an easy-to-use software tool to aid users in
planning their experiments. By quickly making an experimental plan that will be
tailored to the needs of the scientist (enough peaks measured for a structure
analysis), but without spending time collecting redundant data, CrystalPlan will
greatly increase the scientific productivity and throughput of TOPAZ and other
beamlines. Users will be able to measure samples highly efficiently, allowing
them to explore parameters such as temperature and pressure more fully.

@@

\begin{itemize}
  \item Load an instrument geometry file.
  \item Calculate the orientation of a crystal at different goniometer settings.
  \item Calculate the coverage of these orientations in reciprocal space,
  in volume and for single peaks.
  \item Take crystal Bravais lattice symmetries and the higher redundancy of
  higher space groups into account.
  \item Optimize the experiment plan automatically.
  \item Compare predicted and actual measurement of peaks.
\end{itemize}

@@

CrystalPlan currently supports area detectors with flat faces and arbitrary
rectangular dimensions; this reflects the detector types in use at SNS (Anger
cameras and 8-pack $^3$He tube detectors), but the software could easily be
extended for more complicated detector shapes as needed. The detector configuration is loaded from
a text file containing the location, size, orientation and number of pixels of
each detector. Other settings to be adjusted for an instrument include the
detectable range of neutron/x-ray wavelength.

@@

CrystalPlan can accomodate a variety of sample
orientation goniometers. A base Goniometer class is subclassed for specific types of
goniometers, allowing a unified interface. This base class acts as a
\emph{universal goniometer},
which uses the conventional Eulerian goniometer angles of
$\phi$, $\chi$, $\omega$ as defined in \cite{busing67}.
Each class inheriting from the base Goniometer can convert
the desired $\phi$, $\chi$, $\omega$ angles to the internal motor positions it requires. 
The class hierarchy allows for goniometer limitations: for example, TOPAZ is
currently equipped with a goniometer where $\chi$ is fixed at 45 degrees.
CrystalPlan handles this by limiting the sample orientation angles accessible to the user to $\phi$ and
$\omega$.

The sample unit cell and orientation is typically determined by the data
collection/reduction software. In our case, from an initial data set, the data
analysis program ISAW (Integrated Scattering Analysis
Workbench \cite{Mikkelson05}) provides us with the unit cell of the sample (B
matrix) and the sample mounting orientation (U matrix).
The combined UB matrix is then saved to a
text file along with the lattice parameters. 
CrystalPlan can load this UB matrix file and uses it for reciprocal coverage
calculations and optimization.

@@

The reciprocal space to be measured is divided into a cubic 3D matrix 
of {\bf Q}-space (with {\bf Q} defined as $\frac{2\pi}{d}$). 
The user specifies the minimum d-spacing ($d_{min}$, in \ang) 
that he or she is interested in, defining the limit
$Q_{max} = \frac{2\pi}{d_{min}}$ of the modeled volume. There is no lower limit
on {\bf Q}.

The resolution of the 3D matrix is also specified
(in \ang$^{-1}$), which determines the number of points (voxels or volumetric pixels) to be modeled. Available machine memory
limits the resolution that can be achieved in practice.

Each point of the reciprocal space 3D matrix holds an integer representing the 
number of times that voxel of {\bf Q}-space has
been measured -- this is the total coverage matrix. To speed up the
calculation of the total, a coverage matrix for each sample orientation is 
calculated and saved in memory and the total coverage
matrix can be computed by simply adding multiple 3D matrices corresponding to each sample orientation.

@@

For each sample orientation, a rotation matrix combining the sample's mounting
orientation (the U matrix loaded from ISAW) and the goniometer rotation matrices
($\Phi$, $\chi$, $\Omega$) is computed. The coverage in reciprocal space is
computed in this way: for each pixel in the face of a detector, the direction of
the scattered beam is known. Then, two incident beam vectors are considered,
corresponding to the $\lambda_{min}$ and $\lambda_{max}$ wavelength limits
of the detectors and/or source. The scattered beam direction and the two possible incident beam vectors
are used to find two $\overrightarrow q$ vectors; these point in the
same direction. At the high wavelength limit we find $\overrightarrow q_{min}$,
and the low wavelength limit is scattering off of $\overrightarrow q_{max}$.
Since the instrument has a wavelength bandwidth, we fill in the 3D {\bf Q}-space
matrix by filling all points from $\overrightarrow q_{min}$ to the 
$\overrightarrow q_{max}$, for all pixels in the face of the detector.
A typical range for TOPAZ is 0.5 to 3.6 \ang, a 3 \ang\ wavelength band.
This readily defines a volume of coverage, as shown schematically in
two dimensions in Figure \ref{fig:ewald_fig}.

To reduce memory usage yet allow for quickly trying different detector
configurations, the 3D matrix of each sample orientation is saved as a 32- or
64-bit integer, with each bit representing a particular detector. A simple
binary mask can then be applied to quickly compute the coverage if some
detectors are disabled, for example.

@@

Each $hkl$ peak is a separate object that holds all predicted
recordings, including which detector measured it, at what (x,y) position on the
detector face, at what wavelength, and for which sample orientation. These data
is used to  graphically display coverage and calculate statistics such as
redundancy.

\subsection{Crystal Symmetry}
\label{sec:symm}
The crystal's point group symmetry can
be entered along with its lattice parameters.
This information is then used to
calculate the increased measurement redundancy for higher symmetries
\cite{giacovazzo92}. 
A list of $n$ $3\times3$ multiplication matrices is generated based on the point group; 
all $hkl$ vectors are multiplied by
these matrices to find the $n$ equivalent $hkl$ reflections; for example, for
orthorombic (mmm) symmetry, $n=8$; therefore 8 multiplication matrices are
generated that multiply $hkl$ to give $\pm h$, $\pm k$, $\pm l$.

Each reflection is tracked separately, but a primary
$hkl$ is identified that points to a list of equivalent $hkl$'s when the
crystal symmetry is known.
The interactive GUI displays either all $hkl$, ignoring crystal symmetry; or
combines them to display only the primary $hkl$'s, and calculates the
redundancy by summing all measurements on all equivalent $hkl$'s.

In volume coverage mode, a 4D array is generated where the first 3 dimensions
correspond to {\bf Q}-space, and the $4^{th}$ dimension is a list of $n$
entries. Each entry in the $4^{th}$ dimension points to a voxel that is an
equivalent due to crystal symmetry. This allows CrystalPlan to quickly compute a volume coverage
map that takes into account crystal symmetry and the resulting redundancy: at
each voxel, all $n$ equivalent voxels are checked to see if any of them were
measured.

@@

CrystalPlan allows the user to manually enter lists of sample orientations to
simulate and graphically display the coverage in reciprocal space;
however, finding an optimal plan by trial and error is in general very
time consuming and only feasible for an experienced crytallographer.
To make the process more efficient, CrystalPlan includes an Automatic
Coverage Optimizer, which takes as its inputs the desired number of sample
orientations and uses a genetic algorithm to find a solution that satisfies a
user-specified coverage criterion (for example, the user may ask for $85\%$ of 
Bragg reflections to be measured, or for $90\%$ of {\bf Q}-volume).

Genetic algorithms have been covered in the literature before, see e.g.
\cite{Goldberg89}. In this application, the genes
consist of goniometer angles. For instance, if a goniometer has freedom in 3 angles and the user
requests 10 different sample orientations, then each individual has 30 genes.
The initial population is created randomly and is made to evolve using mutations
and crossover. Individuals whose goniometer angles turn out to be impossible to
reach or beyond set limits are eliminated. For each individual set of sample
orientations, the coverage statistics are computed (taking crystal symmetry into
account if desired), and this is used as the fitness of the given individual in
the normal genetic algorithm process. The best individuals are saved after each
cycle. The simulation has succeeded once the desired value is reached or
exceeded or the maximum number of cycles was calculated.

@@

The program was designed to be cross-platform, and run equally well on Linux,
Mac OS and Windows, although Linux is our primary target operating system. Another
design goal was to use of open-source software and toolkits as much as possible.

To this end, CrystalPlan was written in Python \cite{python}, a powerful, open,
cross-platform scripting language. Python was chosen for its ease of cross-platform deployment, and for
its attractive language features. The numpy and scipy libraries (two
open-source, commonly installed Python libraries) were used for many of the
calculations \cite{numpy,scipy}. The most time-critical calculations
were also written with inline C code, which increases the computation speed by
up to $400\times$ as compared with the pure Python version (which can still be
used, in case of incompatibilities).

@@

The calculation code of CrystalPlan is placed in a model layer and kept
completely separate from eventual GUI elements, using a Model/View-Controller
architecture \cite{Gamma95}. This makes it possible to run calculations using
Python scripts, if desired. Unit tests for each part of the calculations are used to ensure
consistent results on different platforms and  systems.

@@

An attractive and easy-to-use GUI was considered an important part of
CrystalPlan, since the program is meant to be used interactively as a guide to
the experimenter whether new or experienced. This requires intuitive GUI
interfaces and workflows.
The user interface was written using wxPython
\cite{wxPython}, a mature cross-platform GUI toolkit, as well as the Enthought
Traits GUI. Some of the 2D plots use matplotlib \cite{matplotlib},  and the 3D
visualization elements use the Mayavi2 toolkit for Python \cite{mayavi}, itself
based on VTK \cite{VTK}. All of these libraries are  open-source.

@@

\begin{itemize}
  	\item {\bf Q}-Space tab: define the volume of interest (minimum d-spacing)
  	and resolution of reciprocal space. 
 	\item Detectors tab: load instrument detector geometry
files, and enable or disable individual detectors. 
	\item Goniometer tab: to choose the
goniometer and degrees of orientation freedom of the sample.
  	\item Sample tab: enter or load the crystal's lattice parameters and UB
  matrix.
  	\item Try an Orientation tab: interactively rotate a sample and observe the
  changes in coverage in real-time.
  	\item Add Orientations tab: type in a list of sample orientation angles to
  add to the experiment plan.

  	\item Experiment Plan tab: a list of the previously calculated sample
orientations. Here you can manage the orientations you will use in the experiment: add,
delete, or temporarily disable them to see the effect on coverage. Once
complete, the plan can be sent to the SNS/TOPAZ data acquisition computers in
order to begin a run \cite{SNSDAS}.

@@

The volume coverage shows a solid isosurface representing the volume measured at
least once. Redundant regions can also be shown using semi-transparent
isosurfaces; or the coverage can be inverted, showing non-measured volumes only;
or slices between user-selectable $q_{min}$ and $q_{max}$.

@@

Both views can take into account crystal symmetry in their display using the
techniques described previously in Section \ref{sec:symm}.
The volume coverage (Figures \ref{fig:volume_view}b) fills in the sphere of
coverage; 
whereas the single reflection view (Figure \ref{fig:refl_view}b) leaves only
the primary reflections (one eight of the total, in this example).

Finally, both views display coverage statistics in the lower right
corner, showing the percentage of volume (or of the number of peaks) measured,
as well as the proportion of redundancy (voxels or peaks measured 2 or more
times).

\subsubsection{Single Reflection View.}
Figure \ref{fig:single_reflection} shows a screenshot of the GUI allowing one to
observe in detail the predicted measurements of a particular peak, by clicking
in the 3D view or by typing in the $hkl$ values. The position on the detector
and wavelength of measurement are displayed. An additional window (not shown) allows the user to place the peak on a particular
position on a detector by automatically finding the necessary sample orientation
angles. This is used, for example, to center important peaks or diffraction
directions in the most sensitive part of a detector.

@@

The resulting UB matrix was loaded into CrystalPlan, where the Automatic
Coverage Optimizer was used to generate an experiment plan with 8 sample
orientations (Figure \ref{fig:natrolite_gui}a) that covered 97\% of the peaks,
taking into account the crystal Laue symmetry (orthorombic or mmm) (Figure
\ref{fig:natrolite_gui}b). The eight orientations were run
unattended overnight.

@@

One planned future development for CrystalPlan include extending the software
to help predict coverage in inelastic neutron scattering experiments. Another
desired feature is the introduction of feedback loops to help determine the
right time to end data acquisition at a particular sample orientation.
CrystalPlan would interface with data analysis software such as ISAW and use
recently developped capabilities for real-time data analysis to monitor
the  signal to noise ratio on particular peaks. When the signal intensity
becomes sufficient, the sample will automatically be moved to the next sample
orientation, further adjusting data acquisition time for each specific sample.
Combined with advanced sample changers coming on-line at TOPAZ, this will allow
for extremely efficient, unattended and automated data acquisition.
